Gut-resident microbial species and their genes are associated with cognition and neuroanatomy in children

### 2023-03-29 manuscript/main.tex
fix colors in table 2

\begin{sidewaysfigure}[!htb]
    \centering
    \includegraphics[width=0.9\textwidth]{assets/Figure1.png}
    \caption{
        \label{fig:1} Study design and overview. (\textbf{A}) Stool samples,
        cognitive assessments and neuroimaging were collected from participants
        at different ages throughout the first years of life.
        (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized using full-scale composite scores.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }

@@

\begin{table}[!htb]
    \centering
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{Group} & \textbf{Subgroup} & \textbf{All} & \textbf{Under 6 months} & \textbf{Over 18 months} \\\hline
      N subjects &  & 361 & 74 & 257 \\  \hline
    \multirow{3}{*}{Samples} & 1 & 262 (72.6\%) & 74 (100.0\%) & 203 (79.0\%) \\ \cline{2-5}
         & 2 & 72 (19.9\%) & 0 (0.0\%) & 46 (17.9\%) \\ \cline{2-5}
         & \textgreater 2 & 27 (7.5\%) & 0 (0.0\%) & 8 (3.1\%) \\ \hline
    \multirow{3}{*}{Age (months)} & min & 2.76 & 2.76 & 18.03 \\ \cline{2-5}
         & max & 119.3 & 6.0 & 119.3 \\ \cline{2-5}
         & median & 25.79 & 3.66 & 45.92 \\ \hline
    \multirow{2}{*}{Sex} & F & 163 (45.2\%) & 37 (50.0\%) & 114 (44.4\%) \\   \cline{2-5}
                         & M & 198 (54.8\%) & 37 (50.0\%) & 143 (55.6\%) \\  \hline
    \multirow{6}{*}{Race} & White & 252 (69.8\%) & 42 (56.8\%) & 191 (74.3\%) \\   \cline{2-5}
        & Black & 40 (11.1\%) & 16 (21.6\%) & 21 (8.2\%) \\ \cline{2-5}
        & Indiginous & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\ \cline{2-5}
        & Asian & 7 (1.9\%) & 1 (1.4\%) & 6 (2.3\%) \\ \cline{2-5}
        & Mixed & 48 (13.3\%) & 12 (16.2\%) & 29 (11.3\%) \\ \cline{2-5}
        & Other & 2 (0.6\%) & 1 (1.4\%) & 2 (0.8\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\   \cline{2-5}
        & Some high school & 6 (1.7\%) & 3 (4.1\%) & 4 (1.6\%) \\ \cline{2-5}
        & High school grad & 37 (10.2\%) & 13 (17.6\%) & 16 (6.2\%) \\ \cline{2-5}
        & Some college & 95 (26.3\%) & 25 (33.8\%) & 57 (22.2\%) \\ \cline{2-5}
        & College grad & 90 (24.9\%) & 17 (23.0\%) & 69 (26.8\%) \\ \cline{2-5}
        & Grad/professional school & 123 (34.1\%) & 15 (20.3\%) & 104 (40.5\%) \\\hline\hline
    \end{tabular}
    \caption{\label{tab:demo}Subject demographics of the ECHO RESONANCE cohort.}
\end{table}

@@

\begin{table}[!htb]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|r|r|}
      \hline\hline
      \textbf{Age group} & \multicolumn{2}{c}{\textbf{0 to 120 months}} & \multicolumn{2}{c}{\textbf{0 to 6 months}} & \multicolumn{2}{c}{\textbf{18 to 120 months}} \\ \hline
      geneset & E.S. & q value & E.S. & q value & E.S. & q value \\\hline
      Acetate syn. I & -0.122 & 0.555 & -0.194 & \color{red}{\textbf{0.153}} & 0.127 & 0.602 \\
      Acetate syn. II & -0.22 & \color{red}{\textbf{0.066}} & -0.342 & \color{red}{\textbf{0.02}} & -0.095 & 0.92 \\
      Acetate syn. III & -0.21 & \color{red}{\textbf{0.086}} & -0.31 & \color{red}{\textbf{0.052}} & -0.086 & 0.92 \\
      Glutamate deg. I & -0.404 & \color{red}{\textbf{0.18}} & -0.284 & 0.816 & -0.249 & 0.602 \\
      Glutamate syn. II & 0.095 & 0.602 & -0.134 & 0.98 & 0.242 & \color{blue}{\textbf{0.047}} \\
      Menaquinone syn. I & -0.085 & 0.563 & -0.083 & 0.816 & -0.17 & \color{red}{\textbf{0.182}} \\
      Propionate deg. I & -0.461 & \color{red}{\textbf{0.13}} & -0.542 & \color{red}{\textbf{0.02}} & -0.674 & \color{red}{\textbf{0.041}} \\
      Propionate syn. I & -0.334 & 0.555 & -0.32 & 0.571 & -0.676 & \color{red}{\textbf{0.023}} \\
      Tryptophan syn. & 0.057 & 0.974 & -0.072 & 0.816 & 0.119 & \color{blue}{\textbf{0.041}} \\\hline\hline
    \end{tabular}
    \caption{\label{tab:fsea}Feature set enrichment analysis on neuroactive microbial gene sets}
    \end{center}
\end{table}

>>>>

\begin{sidewaysfigure}[!h]
    \centering
    \includegraphics[width=0.9\textwidth]{assets/Figure1.png}
    \caption{
        \label{fig:1} Study design and overview. (\textbf{A}) Stool samples,
        cognitive assessments and neuroimaging were collected from participants
        at different ages throughout the first years of life.
        (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized using full-scale composite scores.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }

@@

\begin{table}[!h]
    \centering
    \begin{tabular}{|r|r|r|r|r|}
      \hline\hline
      \textbf{Group} & \textbf{Subgroup} & \textbf{All} & \textbf{Under 6 months} & \textbf{Over 18 months} \\\hline
      N subjects &  & 361 & 74 & 257 \\  \hline
    \multirow{3}{*}{Samples} & 1 & 262 (72.6\%) & 74 (100.0\%) & 203 (79.0\%) \\ \cline{2-5}
         & 2 & 72 (19.9\%) & 0 (0.0\%) & 46 (17.9\%) \\ \cline{2-5}
         & \textgreater 2 & 27 (7.5\%) & 0 (0.0\%) & 8 (3.1\%) \\ \hline
    \multirow{3}{*}{Age (months)} & min & 2.76 & 2.76 & 18.03 \\ \cline{2-5}
         & max & 119.3 & 6.0 & 119.3 \\ \cline{2-5}
         & median & 25.79 & 3.66 & 45.92 \\ \hline
    \multirow{2}{*}{Sex} & F & 163 (45.2\%) & 37 (50.0\%) & 114 (44.4\%) \\   \cline{2-5}
                         & M & 198 (54.8\%) & 37 (50.0\%) & 143 (55.6\%) \\  \hline
    \multirow{6}{*}{Race} & White & 252 (69.8\%) & 42 (56.8\%) & 191 (74.3\%) \\   \cline{2-5}
        & Black & 40 (11.1\%) & 16 (21.6\%) & 21 (8.2\%) \\ \cline{2-5}
        & Indiginous & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\ \cline{2-5}
        & Asian & 7 (1.9\%) & 1 (1.4\%) & 6 (2.3\%) \\ \cline{2-5}
        & Mixed & 48 (13.3\%) & 12 (16.2\%) & 29 (11.3\%) \\ \cline{2-5}
        & Other & 2 (0.6\%) & 1 (1.4\%) & 2 (0.8\%) \\ \hline
    \multirow{6}{*}{Maternal Ed.} & Junior high school & 1 (0.3\%) & 0 (0.0\%) & 1 (0.4\%) \\   \cline{2-5}
        & Some high school & 6 (1.7\%) & 3 (4.1\%) & 4 (1.6\%) \\ \cline{2-5}
        & High school grad & 37 (10.2\%) & 13 (17.6\%) & 16 (6.2\%) \\ \cline{2-5}
        & Some college & 95 (26.3\%) & 25 (33.8\%) & 57 (22.2\%) \\ \cline{2-5}
        & College grad & 90 (24.9\%) & 17 (23.0\%) & 69 (26.8\%) \\ \cline{2-5}
        & Grad/professional school & 123 (34.1\%) & 15 (20.3\%) & 104 (40.5\%) \\\hline\hline
    \end{tabular}
    \caption{\label{tab:demo}Subject demographics of the ECHO RESONANCE cohort.}
\end{table}

@@

\begin{table}[!h]
    \begin{center}
    \begin{tabular}{|r|r|r|r|r|r|r|}
      \hline\hline
      \textbf{Age group} & \multicolumn{2}{c}{\textbf{0 to 120 months}} & \multicolumn{2}{c}{\textbf{0 to 6 months}} & \multicolumn{2}{c}{\textbf{18 to 120 months}} \\ \hline
      geneset & E.S. & q value & E.S. & q value & E.S. & q value \\\hline
      Acetate syn. I & -0.122 & 0.555 & -0.194 & \color{blue}{\textbf{0.153}} & 0.127 & 0.602 \\
      Acetate syn. II & -0.22 & \color{blue}{\textbf{0.066}} & -0.342 & \color{blue}{\textbf{0.02}} & -0.095 & 0.92 \\
      Acetate syn. III & -0.21 & \color{blue}{\textbf{0.086}} & -0.31 & \color{blue}{\textbf{0.052}} & -0.086 & 0.92 \\
      Glutamate deg. I & -0.404 & \color{blue}{\textbf{0.18}} & -0.284 & 0.816 & -0.249 & 0.602 \\
      Glutamate syn. II & 0.095 & 0.602 & -0.134 & 0.98 & 0.242 & \color{red}{\textbf{0.047}} \\
      Menaquinone syn. I & -0.085 & 0.563 & -0.083 & 0.816 & -0.17 & \color{blue}{\textbf{0.182}} \\
      Propionate deg. I & -0.461 & \color{blue}{\textbf{0.13}} & -0.542 & \color{blue}{\textbf{0.02}} & -0.674 & \color{blue}{\textbf{0.041}} \\
      Propionate syn. I & -0.334 & 0.555 & -0.32 & 0.571 & -0.676 & \color{blue}{\textbf{0.023}} \\
      Tryptophan syn. & 0.057 & 0.974 & -0.072 & 0.816 & 0.119 & \color{red}{\textbf{0.041}} \\\hline\hline
    \end{tabular}
    \caption{\label{tab:fsea}Feature set enrichment analysis on neuroactive microbial gene sets.
    Red and blue text indicates significant (q value \textless 0.2) genesets with a positive or negative
    pseudomedian respectively.}
    \end{center}
\end{table}
